Finiteness in Cubical Type Theory

### 2020-01-18 paper.tex
more

\input{preamble}
\title{Finiteness in Cubical Type Theory}
\author{Donnacha Oisín Kidney}
\authorrunning{D. O. Kidney}
\institute{University College Cork \and
  \email{o.kidney@cs.ucc.ie}}
\author{Gregory Provan}
\institute{University College Cork \and
  \email{g.provan@cs.ucc.ie}}
\begin{document}
\maketitle
\begin{abstract}
  We study five different notions of finiteness in Cubical Type Theory and prove
  the relationship between them.
  In particular we show that any totally ordered Kuratowski finite type is
  manifestly Bishop finite.

@@

Finally, the other important property of constructive mathematics is that we can
actually \emph{compute} with proofs.
Cubical Type Theory \cite{cohenCubicalTypeTheory2016}, and its implementation in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, realise this property
even in the presence of univalence, giving computational content to programs
written in HoTT.
\subsection{Contributions}
In this work we will examine five notions of finiteness in Homotopy Type Theory:
the relationships between them, and their topos-theoretic characterisation.
We also briefly examine a definition for countable sets, comparing it to the
manifestly Bishop finite sets.
Our work is formalised in Cubical Agda, where we also develop a library for
proof search based on the finiteness predicates.

@@

The finiteness predicates we are interested in are organised in
Figure~\ref{finite-classification}.
We will provide a definition for each, and prove each relation to the other
finiteness predicates.
In terms of Figure~\ref{finite-classification}, this means providing a function
to inhabit each arrows.

@@

\begin{itemize}
  \item Relation between different versions of finiteness, especially card to
    bishop.
  \item Topos for Kuratowski
  \item Countable and W-topos
  \item Search
\end{itemize}
\subsection{Related Work}
\subsection{Notation}

% Taking finiteness as an example again, if we prove that some type is finite via
% bijection, in MLTT such a proof would have to reveal the particular bijection.
% This becomes problematic when defining, for instance, the category of finite
% sets: the objects of this category would ideally be types coupled with proofs of
% their finiteness, but in order to

% The five in particular that we are interested in are organised in
% figure~\ref{finite-classification}.
% We especially want to examine what separates each predicate: what extra
% information do you need, for instance, to go from a proof of Cardinal finiteness
% (section~\ref{cardinal}) to manifest Bishop finiteness
% (section~\ref{manifest-bishop}).

% Next, we will examine the category-theoretic and topos-theoretic interpretations
% of each of the predicates.
% We will give a ``category of finite sets'' for each finiteness predicate, and
% then prove the various closure properties required to arrive at the relevant
% topos (we show that decidable Kuratowski finite sets form a \(\Pi\)-pretopos,
% for instance).

% ways to prove finiteness end up proving the same
% thing: constructively, however, each defines a fundamentally different
% ``version'' of finiteness.
% In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
% variations (and more) have been explored in
% \cite{spiwackConstructivelyFinite2010} and
% \cite{firsovVariationsNoetherianness2016}.

% Homotopy Type Theory \cite{hottbook} further adds to this story in two ways.
% First, its universe of types includes non-sets: whether or not a finiteness
% predicate includes these gives us another dimension on which to discriminate.

% The second addition is concerned with the hiding of information.
% Occasionally in a constructive setting you don't want to have to reveal how you
% proved something.
% To model logical disjunction, for instance, we usually use a disjoint union: the
% model isn't entirely correct, though, because a proof of \(A \text{or} B\)
% reveals which of \(A\) or \(B\) we actually proved.

% logical disjunction, for instance, is usually modelled with a
% disjoint union.
% This always means, however, that to prove ``A or B'' you have to reveal which of
% A or B you did prove.

% The other headline feature of HoTT's generalised equality is univalence: put
% simply, it allows us to treat isomorphic types as equal, giving us in the
% constructive world access to a technique that is commonplace classically.
% It turns out that finite types have a lot to say about isomorphisms, and we will
% rely heavily on this aspect of HoTT to work with them.
% \subsection{Contributions}
% In this work, we will explore finite types in Cubical Type Theory
% \cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
% types, species, and demonstrate their practical uses for proofs in dependently
% typed programming languages.

% In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
% finiteness (i.e. those at least as strong as Kuratowski finiteness
% \cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
% finiteness.
% We will delineate some subtly different finiteness predicates, and prove the
% relationship between each.
% We will prove, then, that every discrete Kuratowski finite type is cardinal
% finite, and subsequently will show that discrete Kuratowski finite sets form a
% topos.

% In section~\ref{species-and-containers}, we will redefine our finiteness
% predicates using \emph{containers}, and prove that the new definitions are
% equivalent to the old.
% Using this, we will prove much stronger relationships between our finiteness
% predicates and relation-based definitions of finiteness.
% Finally, We will show how the species-container duality falls out naturally from
% these new definitions.

% In section~\ref{infinite-cardinalities}, we will extend our study of finite
% types to infinite but countable types.
% We will see that the finiteness predicates are mirrored with countable
% counterparts, and we will prove closure under the Kleene star and plus.

% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.

% Classically, of course, each of these statements are equivalent.
% Constructively, though, all four predicates are fundamentally different.

% Part of the difference comes from ``leaking'' information about the finite type
% in the proof of its finiteness.
% If I demonstrate a type is finite by showing a surjection to that type, I've
% revealed a way to construct all of the elements of my type.
% If I show that it is bounded, I reveal an upper bound on the cardinality of the
% type.

\subsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

Figure~\ref{finite-classification} organises the predicates according to
their strength.
We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.
Our main proof in this regard is that we can derive manifest Bishop finiteness
from Kuratowski finiteness plus a total order.

We will---through the use of containers
\cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
these predicates have with the usual function relations i.e. we will show that a
proof of manifest enumerability is precisely equivalent to a surjection from a
finite prefix of the natural numbers.

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos.
\subsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the classification of finiteness predicates mirrors that of
their countable counterparts, and we will prove closure under the Kleene star.

\subsection{Practical Uses}
Finally, in section~\ref{practical} we will demonstrate some practical uses of
the finiteness proofs in Cubical Agda.
We will show how to use well-known techniques
\cite{firsovDependentlyTypedProgramming2015, devrieseBrightSideType2011} to
automate proofs of functions with finite (and infinite) domains.
We further show how to automate the synthesis of functions from desired
properties.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.

Before giving the definition of the predicate, we will first define lists (and
membership thereof): we have chosen a \emph{container}-based definition for this
work.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket(A) = \Sigma {(s : S)} , \left( P(s) \rightarrow A \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
  Where \(\text{fiber}\) is from \cite[definition 4.2.4]{hottbook}.
\end{rm-definition}
\begin{rm-definition}[Lists]
  \begin{equation}%
    \mathbf{List} = \llbracket \mathbb{N} \triangleright \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\).
  We define it the standard way, where \(\mathbf{Fin}(0) = \bot\) and
  \(\mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)\).
\end{rm-definition}

Internally, in our formalisation, we actually use the standard
inductive definition of lists more often (it tends to work better in more
complex algorithms, and functions on it seems to satisfy the termination
checker more readily).
However, since both types are equivalent, univalence allows us to transport to
whichever representation is more convenient in a given situation.
For the higher-level proofs we present here, though, the container-based
definition greatly simplifies certain steps, which is why we have chosen it as
our representation.

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  This definition is what is referred to as \verb+Listable+ in
\end{rm-definition}
\cite{firsovDependentlyTypedProgramming2015}.
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
    \begin{equation}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
    \begin{equation}
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
    \end{equation}
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
\end{rm-definition}
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \iff \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  After sufficient inlining, it emerges that our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma} \label{split-enum-bool-top-bot}
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  Each of these types clearly has a finite number of elements (2, 1, and 0,
  respectively), and furthermore has a straightforward enumeration.
\end{proof}
\begin{rm-theorem} \label{split-enum-sigma}
  Split-enumerability is closed under \(\Sigma\).
  \begin{equation} \label{split-enum-sigma-clos-sig}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-sup}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction} \cite[definition
  3.11.1]{hottbook}; i.e. its membership proof exists, and all such proofs are
  equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}
By the definition of a contraction, we can always recover the underlying
membership proof, meaning that we can always derive split enumerability from
manifest Bishop finiteness.
\subsection{Equivalence}
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite[definition 4.4.1]{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function (called \(\text{uniques}\)) which filters
  out duplicates from lists of \(A\)s.
  This gives us our support list.
  To generate the cover proof it suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
\end{proof}
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

>>>>

\input{preamble}
\title{Finiteness in Cubical Type Theory}
\author{
  Donnacha Oisín Kidney \inst{1} \and
  Gregory Provan \inst{2}
}
\institute{
  University College Cork \email{o.kidney@cs.ucc.ie} \and
  University College Cork \email{g.provan@cs.ucc.ie}
}
\begin{document}
\maketitle
\begin{abstract}
  We study five different notions of finiteness in Cubical Type Theory and prove
  the relationship between them.
  In particular we show that any totally ordered Kuratowski finite type is
  manifestly Bishop finite.

@@

Finally, the other important property of constructive mathematics is that we can
actually \emph{compute} with proofs.
Cubical Type Theory \cite{cohenCubicalTypeTheory2016}, and its implementation in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, realise this property
even in the presence of univalence, giving computational content to programs
written in HoTT.
\subsection{Contributions}
In this work we will examine five notions of finiteness in Homotopy Type Theory,
the relationships between them, and their topos-theoretic characterisation.
We also briefly examine a definition for countable sets, comparing it to the
manifestly Bishop finite sets.
Our work is formalised in Cubical Agda, where we also develop a library for
proof search based on the finiteness predicates.

@@

The finiteness predicates we are interested in are organised in
Figure~\ref{finite-classification}.
We will explore two aspects of each predicate: its relation with the other
predicates, and its topos-theoretic classification.

When we say ``relation'' we are referring to the arrows in
Figure~\ref{finite-classification}.
The proofs, then, amount to a function which inhabits each arrow.
Each unlabelled arrow is a weakening: i.e., every manifest Bishop finite set
is manifest enumerable.
The labelled arrows are strengthening proofs: i.e. every manifest enumerable set
\emph{with decidable equality} is manifest Bishop finite.

@@

\todo[inline]{redo these last two paragraphs}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates.

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
We also use this section to demonstrate some \emph{computation} using
univalence: our proof search library can deal seamlessly with functions, testing
them for equality, synthesising them, etc.
\subsection{Related Work}
\subsection{Notation}

\section{Finiteness Predicates}
\subsection{Split Enumerable}
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  This definition is what is referred to as \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{rm-definition}

\section{Relations Between Each Finiteness Definition}
\section{Topos}
\section{Countably Infinite Types}
\section{Search}

% % Taking finiteness as an example again, if we prove that some type is finite via
% % bijection, in MLTT such a proof would have to reveal the particular bijection.
% % This becomes problematic when defining, for instance, the category of finite
% % sets: the objects of this category would ideally be types coupled with proofs of
% % their finiteness, but in order to

% % The five in particular that we are interested in are organised in
% % figure~\ref{finite-classification}.
% % We especially want to examine what separates each predicate: what extra
% % information do you need, for instance, to go from a proof of Cardinal finiteness
% % (section~\ref{cardinal}) to manifest Bishop finiteness
% % (section~\ref{manifest-bishop}).

% % Next, we will examine the category-theoretic and topos-theoretic interpretations
% % of each of the predicates.
% % We will give a ``category of finite sets'' for each finiteness predicate, and
% % then prove the various closure properties required to arrive at the relevant
% % topos (we show that decidable Kuratowski finite sets form a \(\Pi\)-pretopos,
% % for instance).

% % ways to prove finiteness end up proving the same
% % thing: constructively, however, each defines a fundamentally different
% % ``version'' of finiteness.
% % In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
% % variations (and more) have been explored in
% % \cite{spiwackConstructivelyFinite2010} and
% % \cite{firsovVariationsNoetherianness2016}.

% % Homotopy Type Theory \cite{hottbook} further adds to this story in two ways.
% % First, its universe of types includes non-sets: whether or not a finiteness
% % predicate includes these gives us another dimension on which to discriminate.

% % The second addition is concerned with the hiding of information.
% % Occasionally in a constructive setting you don't want to have to reveal how you
% % proved something.
% % To model logical disjunction, for instance, we usually use a disjoint union: the
% % model isn't entirely correct, though, because a proof of \(A \text{or} B\)
% % reveals which of \(A\) or \(B\) we actually proved.

% % logical disjunction, for instance, is usually modelled with a
% % disjoint union.
% % This always means, however, that to prove ``A or B'' you have to reveal which of
% % A or B you did prove.

% % The other headline feature of HoTT's generalised equality is univalence: put
% % simply, it allows us to treat isomorphic types as equal, giving us in the
% % constructive world access to a technique that is commonplace classically.
% % It turns out that finite types have a lot to say about isomorphisms, and we will
% % rely heavily on this aspect of HoTT to work with them.
% % \subsection{Contributions}
% % In this work, we will explore finite types in Cubical Type Theory
% % \cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
% % types, species, and demonstrate their practical uses for proofs in dependently
% % typed programming languages.

% % In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
% % finiteness (i.e. those at least as strong as Kuratowski finiteness
% % \cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
% % finiteness.
% % We will delineate some subtly different finiteness predicates, and prove the
% % relationship between each.
% % We will prove, then, that every discrete Kuratowski finite type is cardinal
% % finite, and subsequently will show that discrete Kuratowski finite sets form a
% % topos.

% % In section~\ref{species-and-containers}, we will redefine our finiteness
% % predicates using \emph{containers}, and prove that the new definitions are
% % equivalent to the old.
% % Using this, we will prove much stronger relationships between our finiteness
% % predicates and relation-based definitions of finiteness.
% % Finally, We will show how the species-container duality falls out naturally from
% % these new definitions.

% % In section~\ref{infinite-cardinalities}, we will extend our study of finite
% % types to infinite but countable types.
% % We will see that the finiteness predicates are mirrored with countable
% % counterparts, and we will prove closure under the Kleene star and plus.

% % Proofs of finiteness have well-known practical applications in
% % constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% % In section~\ref{practical}, we build a library which exploits these uses in
% % Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% % complex proofs over finite types.
% % We frame this in terms of the principle of omniscience for finite types.
% % Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% % go further than the usual examples of this kind of proof automation: as well as
% % proving properties about functions, we can synthesise functions whole-cloth from
% % their desired properties.
% % Through the unified interface for finite and countable types, we can
% % reuse the automation machinery for \emph{partial} proof search over infinite
% % search spaces.
% % Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% % prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% % dependent functions.

% % Classically, of course, each of these statements are equivalent.
% % Constructively, though, all four predicates are fundamentally different.

% % Part of the difference comes from ``leaking'' information about the finite type
% % in the proof of its finiteness.
% % If I demonstrate a type is finite by showing a surjection to that type, I've
% % revealed a way to construct all of the elements of my type.
% % If I show that it is bounded, I reveal an upper bound on the cardinality of the
% % type.

% \subsection{Strong Finiteness Predicates}
% We will first explore the ``strong'' notions of finiteness (i.e. those at least
% as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
% with a special focus on cardinal finiteness (section~\ref{cardinal}), and
% manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
% our knowledge.

% Figure~\ref{finite-classification} organises the predicates according to
% their strength.
% We will go through each of the predicates, proving how to weaken each (i.e. we
% will provide a proof that every cardinally finite type is Kuratowski finite),
% and how to strengthen them, given the required property.
% In terms of figure~\ref{finite-classification}, this amounts to providing proofs
% for each arrow.
% Our main proof in this regard is that we can derive manifest Bishop finiteness
% from Kuratowski finiteness plus a total order.

% We will---through the use of containers
% \cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
% these predicates have with the usual function relations i.e. we will show that a
% proof of manifest enumerability is precisely equivalent to a surjection from a
% finite prefix of the natural numbers.

% For each predicate, we will also prove its closure over sums and products in
% both dependent and non-dependent forms, if such a closure exists.
% This will culminate in our main result for this section: the formal proof that
% decidable Kuratowski finite sets form a topos.
% \subsection{Infinite Types}
% In section~\ref{infinite-cardinalities}, we will extend our study of finite
% types to infinite but countable types.
% We will see that the classification of finiteness predicates mirrors that of
% their countable counterparts, and we will prove closure under the Kleene star.

% \subsection{Practical Uses}
% Finally, in section~\ref{practical} we will demonstrate some practical uses of
% the finiteness proofs in Cubical Agda.
% We will show how to use well-known techniques
% \cite{firsovDependentlyTypedProgramming2015, devrieseBrightSideType2011} to
% automate proofs of functions with finite (and infinite) domains.
% We further show how to automate the synthesis of functions from desired
% properties.
% \section{Split Enumerability} \label{split-enumerable}
% We will start with the simplest definition of finiteness: we say a set is
% enumerable if there is a list of its elements which contains every element in
% the set.

% Before giving the definition of the predicate, we will first define lists (and
% membership thereof): we have chosen a \emph{container}-based definition for this
% work.
% \begin{rm-definition}[Container] \label{container-def}
%   A container \cite{abbottContainersConstructingStrictly2005} is a pair
%   \(S \triangleright P\) where \(S\) is a type, the elements of which are called
%   the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
%   the elements of \(P(s)\) are called the \emph{positions} of a container.
%   We ``interpret'' a container into a functor defined like so:
%   \begin{equation}
%     \llbracket S \triangleright P \rrbracket(A) = \Sigma {(s : S)} , \left( P(s) \rightarrow A \right)
%   \end{equation}
%   Membership of a container can be defined like so:
%   \begin{equation}
%     x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
%   \end{equation}
%   Where \(\text{fiber}\) is from \cite[definition 4.2.4]{hottbook}.
% \end{rm-definition}
% \begin{rm-definition}[Lists]
%   \begin{equation}%
%     \mathbf{List} = \llbracket \mathbb{N} \triangleright \mathbf{Fin} \rrbracket
%   \end{equation}
% \end{rm-definition}
% \begin{rm-definition}[\mathbf{Fin}]
%   \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\).
%   We define it the standard way, where \(\mathbf{Fin}(0) = \bot\) and
%   \(\mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)\).
% \end{rm-definition}

% Internally, in our formalisation, we actually use the standard
% inductive definition of lists more often (it tends to work better in more
% complex algorithms, and functions on it seems to satisfy the termination
% checker more readily).
% However, since both types are equivalent, univalence allows us to transport to
% whichever representation is more convenient in a given situation.
% For the higher-level proofs we present here, though, the container-based
% definition greatly simplifies certain steps, which is why we have chosen it as
% our representation.

% Finally, we can define formally split enumerability.
% \subsection{Split Surjections}
% \begin{rm-definition}[Surjections]
%   \begin{multicols}{2}
%     \begin{equation}
%       A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
%     \end{equation} \break
%     \begin{equation}
%       A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
%     \end{equation}
%   \end{multicols}
%   Where the definitions of surjective are taken from \cite[definition
%   4.6.1]{hottbook}.
% \end{rm-definition}
% \begin{rm-theorem} \label{split-enum-surj}
%   Split enumerability is equivalent to a split surjection from a finite prefix
%   of the natural numbers.
%   \begin{equation}
%     \mathcal{E}!(A) \iff \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
%   \end{equation}
% \end{rm-theorem}
% \begin{proof}
%   After sufficient inlining, it emerges that our goal is simply a reassociation.
%   \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-surj}
%   To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
%   have written out the extra lines for clarity alone.
% \end{proof}
% \begin{rm-lemma} \label{split-enum-discrete}
%   Any split enumerable type has decidable equality (is discrete).
% \end{rm-lemma}
% \begin{proof}
%   We use a corollary that if there is a split-surjection from \(A\) to \(B\),
%   and \(A\) is discrete, then \(B\) is also discrete.
% \end{proof}
% By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
% this implies split enumerable types must all be sets.
% \subsection{Closure}
% In this section we will prove closure under various operations for split
% enumerable sets.
% We are working towards a topos proof, which requires us to prove closure under
% a variety of operations: for now, we only have enough machinery to demonstrate
% the semiring operations, and dependent sums.
% In order to show closure under exponentials (function arrows), we will need an
% equivalence with \(\mathbf{Fin}\), which will be provided in
% section~\ref{manifest-bishop}.
% \begin{rm-lemma} \label{split-enum-bool-top-bot}
%   \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
% \end{rm-lemma}
% \begin{proof}
%   Each of these types clearly has a finite number of elements (2, 1, and 0,
%   respectively), and furthermore has a straightforward enumeration.
% \end{proof}
% \begin{rm-theorem} \label{split-enum-sigma}
%   Split-enumerability is closed under \(\Sigma\).
%   \begin{equation} \label{split-enum-sigma-clos-sig}
%     \frac{
%       \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
%     }{
%       \mathcal{E}!(\Sigma (x : A) , U(x))
%     }
%   \end{equation}
% \end{rm-theorem}
% \begin{proof}
%   To obtain the support list, we concatenate the support lists of all the proofs
%   of split-finiteness for \(U\) over the support list of \(E_A\).
%   In Agda:
%   \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-sup}
%   ``do-notation'' is available to us as we're working in the list monad.
% \end{proof}
% Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
% as these types can be defined in terms of \(\Sigma\).
% \begin{align}
%   A \times B &= \Sigma {(x : A)} , B \\
%   A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
% \end{align}
% \section{Manifest Bishop Finiteness} \label{manifest-bishop}
% Where split enumerability was the enumeration form of a split surjection from
% \(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
% \emph{equivalence} with \(\mathbf{Fin}\).
